---
title: "A Comprehensive Analysis of Delay Patterns and Operational Efficiency in the Toronto Transit Commission (TTC) Subway System"
author:
  - Ricky Yuan
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/RickyYuan666/TTC-delay>"
date: 09-26-2024
date-format: long
abstract: "This paper uses publicly available data to explore the patterns and causes of delays in the Toronto Transit Commission (TTC) subway network. Through exploratory data analysis and statistical modeling, the study identifies the most common causes of delays, stations with frequent delays, and the operational impact of these delays on service gaps and vehicle performance. Not only that, this paper also analyzes peak delay times and their correlation with specific subway lines and stations. These findings provide transit operators with solutions to reduce delays and improve the passenger experience. Therefore, this improvement is critical to increasing operational efficiency. This study also provides practical recommendations for reducing future disruptions to the TTC subway system by addressing these issues."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| echo: false
#| warning: false
#### Loading Packages ####
library(tidyverse)
library(opendatatoronto)
```

# Introduction {#sec-introduction}

Public transportation systems are critical to the efficient operation of large urban centers, providing reliable transportation for millions of daily commuters. The Toronto Transit Commission (TTC) subway system is one of the most critical networks in North America and is the primary mode of transportation for the city's residents. However, the Toronto Transit Commission (TTC) subway system suffers from frequent delays that reduce the overall efficiency of the subway system. As the city grows and the demand for efficient public transportation increases, understanding the causes and patterns of these delays is critical to improving efficiency and transit system reliability.

Previous studies have examined the various factors affecting subway delays. For example, @louie2017modelling explored causal (e.g., infrastructure problems) and non-causal (e.g., human error) factors contributing to subway disruptions. Their research emphasizes the complexity of managing the subway system. The reason is multiple interacting factors can prolong delays in a subway system. Additionally, @schleifer2022toronto demonstrated that delay frequency varies significantly by ridership patterns, especially at major transfer points such as Union and Kipling stations. These findings suggest that understanding the specific characteristics of passenger flows and stations is critical to addressing delays within the system.

This paper aims to analyze TTC subway delay data for the year 2023 to reveal patterns of delay occurrence, duration, and the most affected stations and subway lines. This study uses data obtained from OpenDataToronto, visual modeling analysis, and statistical methods to examine trends. Examples include delays over time, the frequency at each station, and the total duration of delays on each subway line. Building on research by @louie2017modelling and @schleifer2022toronto, this analysis focuses on operational inefficiencies and their relationship to specific stations and lines. The analysis results provide actionable insights and recommendations for reducing delays and improving the rider experience within the Toronto subway system.


# Data {#sec-data}

## Overview {#sec-data-overview}

This study uses Toronto Transit Commission (TTC) data from the OpenDataToronto portal for the year 2023. This dataset provides key information on delay events such as the station where the delay occurred, the cause of the delay (e.g., mechanical failure, passenger-related disruptions), the duration of the delay (in minutes), and the time of day. This clear data enables us to distinguish between peak and off-peak hours. These variables allow us to comprehensively understand the patterns and causes of delays on the Toronto subway network.

In addition to the basic structure of the dataset, we constructed several variables during the cleaning process to enhance the analysis. For example, delays were categorized based on their duration, with 0-5 minutes being short, 5-15 minutes being medium, and more than 15 minutes being long. In addition, the dataset was subdivided to differentiate between delays occurring during peak and off-peak hours, which provides a more nuanced understanding of the impact of the time of day when delays occur on the incidence of delays.

This dataset is particularly important for identifying operational inefficiencies within the TTC system. As previous research has shown, causal factors such as infrastructure failures and non-causal factors such as human error can significantly impact subway delay duration (@Louie2017). In addition, ridership levels and station-specific characteristics have also played a crucial role in delays, especially at major transfer points with high concentrations of riders such as Union and Kipling stations (@schleifer2022toronto). By examining these variables, this study aims to gain insight into the most common causes and patterns of delays, which can inform strategies to improve system efficiency.

The analysis employs a variety of visualization and statistical methods to explore temporal, spatial, and route-specific delay patterns. Peak periods of inefficiency are identified by mapping delay durations over time, while spatial visualizations highlight the stations with the most frequent delays. In addition, bar charts can be used to compare the total delay duration of different subway lines, revealing which lines are most prone to failure. By using these methods, the causes of TTC delays can be identified, and solutions can be provided to improve transit system performance.

The analysis results show that the average delay across the network is about 7 minutes. However, delays are significantly longer at Union and Kipling stations, especially during peak hours. This phenomenon has significant scope for operational improvements at specific stations and periods.

The data was cleaned and organized using the janitor [@citejanitor] and tidyverse [@citetidyverse] packages in R [@citeR], and the opendatatoronto [@citeopendatatoronto] package from the OpenDataToronto portal to obtain raw data. After cleaning, the dataset was tested for accuracy and the cleaned dataset was used for further analysis.

## Results {#sec-data-results}

```{r}
#| echo: false
#| warning: false
#| message: false

data <- read_csv(here::here("data","02-analysis_data","analysis_data.csv"))
```

After loading the dataset using the R programming language [@citeR] and the `opendatatoronto` package [@citeopendatatoronto], the `tidyverse` [@citetidyverse] package was used to generate graphs. In doing so, R code was adapted from @tellingstorieswithdata.

## Measurement {#sec-data-measurement}
The dataset used in this study is from the OpenDataToronto portal and focuses on Toronto Transit Commission (TTC) subway delay data for the year 2023. These data are collected by the Toronto Transit Commission (TTC) during its daily operations, using a combination of automated record-keeping systems and manual reporting to track delays across subway lines and stations. Critical variables in the dataset include the station where the delay occurred, the exact date and time of the delay, the type of delay (e.g., "mechanical failure," "signaling problems," "passenger assistance," or "weather conditions"), and the duration of the delay in minutes. Also, station variables are recorded manually or automatically, which can provide spatial information about where the delay occurred. Also, dates and times are recorded using the TTC's internal time tracking system for time analysis. Types of delays are reported by staff or maintenance teams, depending on the event, and the TTC calculates the duration of delays from the time of disruption to the return to regular service using either the automated system or manual recording. These measurements provide a comprehensive understanding of delay patterns in the subway system, documenting when and where delays occur and the causes and durations of delays. Through this robust data collection process, this study can effectively identify spatial and temporal trends in delays.

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-delay-duration-over-time
#| fig-cap: "Average delay duration over time in the TTC subway system from January to August 2024, aggregated monthly."

#### First Graph ####

library(ggplot2)
library(dplyr)
library(lubridate)

# Convert Date to Date format
data$Date <- as.Date(data$Date)

# Aggregate data by month to avoid overcrowding
data$Month <- floor_date(data$Date, "month")
monthly_delay <- data %>%
  group_by(Month) %>%
  summarise(Avg_Delay = mean(`Delay Duration (Minutes)`, na.rm = TRUE))

# Plot: Average delay duration over time (monthly aggregated)
ggplot(monthly_delay, aes(x = Month, y = Avg_Delay)) +
  geom_line(color = "darkgreen", size = 1) +
  labs(
    title = "Average Delay Duration Over Time (Monthly Aggregated)",
    x = "Month",
    y = "Average Delay Duration (Minutes)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-delay-duration-over-time shows the average delay duration (in minutes) for the Toronto Transit Commission (TTC) subway system aggregated by month from January to August 2024. The data shows significant variation in delay duration throughout the year, and the average duration of each delay event ranges from approximately 2.6 minutes to a high of 3.5 minutes.

At the beginning of the year, the average delay duration decreased slightly to under 3 minutes  from January through February. However, delays continued to increase from March through May, peaking at approximately 3.5 minutes in June. The upward trend indicates that the TTC faced significant inefficiencies during this period, which may have affected the overall performance of the subway system. After the June peak, the trend shows a significant decrease in average delay times throughout the summer, reaching a low of about 2.6 minutes in September. However, delays increased slightly in November, possibly due to seasonal factors such as increased ridership or early winter weather disruptions. By December, average delays were again reduced to nearly three minutes, returning to more stable operations.

Overall, the fluctuating delay patterns highlight the need for targeted interventions during periods of high operational stress, especially in the summer and late fall. The TTC can use this data to prioritize maintenance efforts and allocate resources more efficiently, ultimately reducing delays and improving the reliability of subway service throughout the year.

\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-delay-causes-by-station
#| fig-cap: "Top 10 stations with the most delay occurrences in the TTC subway system from January to August 2024."

#### Second Graph ####
# Aggregate data by station and calculate total delays
station_delay <- data %>%
  group_by(Station) %>%
  summarise(Delay_Count = n()) %>%
  arrange(desc(Delay_Count)) %>%
  top_n(10, Delay_Count) # Select top 10 stations with most delays

# Plot: Top 10 stations with most delay counts
ggplot(station_delay, aes(x = reorder(Station, Delay_Count), y = Delay_Count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Top 10 Stations with Most Delay Occurrences",
    x = "Station",
    y = "Count of Delays"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  coord_flip() # Flip to make station names easier to read
```

The @fig-delay-causes-by-station shows the top 10 stations in the Toronto Transit Commission's (TTC) subway system that will experience the highest number of delays from January to August 2024. The bar chart shows that Bloor Station has the highest number of delays at over 800. Kennedy BD Station and Finch Station follow closely behind, with over 700 delays.

These high-traffic stations are such essential transportation components of the system that delays occur frequently. Union Station is a key transfer station for multiple subway lines, ranked seventh with over 500 delays. This result shows that despite its high ridership, Union Station does not experience delays as often as expected. Kipling and Dundas stations, which are essential transfer points or terminals.It is also appear on the list, showing that their operational inefficiencies may stem from train schedules, maintenance challenges, or platform congestion.

The distribution of delays at these top 10 stations highlights that delays are concentrated at major interchanges and terminals. While delays occur throughout the transit system, they are particularly prevalent at stations with high operational complexity and passenger volumes.  To adjust the system, the TTC should optimize resource allocation, maintenance planning, and operations. Therefore, these high-delay stations could significantly improve service reliability and reduce delays throughout the TTC subway system.
\newpage

```{r}
#| echo: false
#| warning: false
#| label: fig-delays-by-line
#| fig-cap: "Total delay duration (in minutes) for each TTC subway line from January to August 2024."

#### Third Graph ####
# Aggregate total delays by line
line_delay <- data %>%
  group_by(Line) %>%
  summarise(Total_Delay = sum(`Delay Duration (Minutes)`, na.rm = TRUE))

# Plot: Total delay by subway line (Bar Chart)
ggplot(line_delay %>% filter(Total_Delay > 60), aes(x = reorder(Line, Total_Delay), y = Total_Delay, fill = Line)) +
  geom_bar(stat = "identity", show.legend = FALSE) + # Hide legend since it's redundant
  labs(
    title = "Total Delay Duration by Subway Line",
    x = "Subway Line",
    y = "Total Delay (Minutes)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for readability
  scale_fill_brewer(palette = "Set3") # Apply a visually appealing color palette
```

@fig-delays-by-line lists the total delays (in minutes) for each subway line in the Toronto Transit Commission (TTC) system from January to August 2024. It is clear from the graph that the Yonge-University line (Line 1) and the Bloor-Danforth line (Line 2) cause the most extended cumulative delays, with Line 1 having the most prolonged delays at over 30,000 minutes, followed closely by Line 2.

In contrast, the Sheppard Line (Line 4) has fewer delays, reflecting its shorter route length and lower ridership than Lines 1 and 2. The differences in delay durations across lines suggest that operational disruptions are more prevalent on the busiest lines, possibly due to higher service demand, frequent train movements, and ridership. These factors contribute to the operational complexity and increased likelihood of delays on Lines 1 and 2.

This data highlights that the TTC has many deficiencies on its main lines. This suggests that focusing on these routes with high levels of delay will yield the most substantial improvements in overall system performance. By prioritizing maintenance and resource allocation on the Yonge-University and Bloor-Danforth lines, the TTC can significantly reduce delays and improve the efficiency and reliability of the subway system.

# Discussion {#sec-discussion}

As shown in @fig-delay-duration-over-time, the TTC subway system has experienced significant fluctuations in delay duration throughout 2023. The most notable peak occurs in June when the average delay duration reaches approximately 3.5 minutes. This shows that operational challenges and external disturbances require more attention during this period. Towards the end of the year, the delays decreased, indicating improved system performance, which could be due to better management practices or reduced external interference.

Additionally, @fig-delay-causes-by-station shows a disproportionate number of delays at Union and Kipling stations, which could be attributed to the high volume of passengers and complex operational demands. As significant hubs in the Metro network, these two stations have a high frequency of delays and, therefore, require targeted interventions to alleviate congestion and reduce operational efficiency.

Finally, @fig-delays-by-line shows that the Bloor-Danforth Line (Line 2) and the Yonge-University Line (Line 1) impact system-wide delays the most. These two lines form the TTC's main routes and connect most of the city's riders. Therefore, given the system-wide impact of these two heavily traveled lines, the TTC should prioritize improving its operations. The analysis in this paper emphasizes the importance of addressing these key stations and lines to reduce overall delays and enhance the rider experience.

\newpage

\appendix

# Appendix {#sec-appendix}

## Dataset and Graph Sketches {#sec-appendix-sketches}

The dataset used in this analysis comes from the OpenDataToronto portal[@citeopendatatoronto], specifically focusing on the Toronto Transit Commission (TTC) subway delay data. This dataset contains detailed records of delay events, including the station, time, date, delay caused, and delay duration. The graphs generated in this analysis include visual representations of delay patterns over time, the frequency of delays by station, and the total delays by subway line, which help provide insights into the operational challenges faced by the TTC in 2023. All visualizations were produced using the ggplot2 library[@citetggplot2] in R[@citeR].

## Data Cleaning {#sec-appendix-cleaning}
The process of data cleaning involved filtering the relevant data for the TTC subway delays. The dataset was cleaned by renaming some columns to make them more readable, removing unnecessary columns, and handling missing values. Therefore, the cleaned dataset is used for analysis, ensuring accuracy and clarity in the visualizations and statistical evaluations.

## Attribution Statement {#sec-appendix-attribution}

This research uses the TTC subway delay data which is licensed under the Open Government Licence of Toronto[@tphlicense].It available at OpenDataToronto[@citeopendatatoronto]. All data and visualizations generated in this analysis are fully compliant with this license. 

\newpage

# References {#sec-references}
